---
title: "TEF4_data_analysis"
author: "DVM Bishop"
date: "9th Feb 2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Source is:  
https://www.officeforstudents.org.uk/advice-and-guidance/teaching/tef-data/get-the-data/

"Each metric shows the proportion of students with successful outcomes. 
This is compared to expected performance for that provider to take into account the mix of students and subjects at each provider. 
If a provider’s actual performance is significantly above its benchmark, this is taken as a measure of high performance." 

Explanation of benchmarking methodology is provided on the 
Higher Education Statistics Agency (HESA) website:
https://www.hesa.ac.uk/data-and-analysis/performance-indicators/benchmarks

See also, re conversion of flags to Gold, Silver, Bronze
https://wonkhe.com/blogs/policy-watch-tef-year-4-data-release/

Contextual data available here under TEF Year Four Contextual Data heading: https://www.officeforstudents. org.uk/advice-and-guidance/teaching/tef-data/get-the-data/

This is another of the many documents purporting to explain TEF metrics, but, as usual, you are just then directed to yet more documents
https://www.officeforstudents.org.uk/.../tef-year-four-metrics-and-contextual-data.pdf

This recommends 'Annex B of the TEF Year Four procedural guidance' for detailed explanation of TEF data.
That document is here:
https://www.officeforstudents.org.uk/publications/teaching-excellence-and-student-outcomes-framework-year-four-procedural-guidance/

There is also this which has several documents all in one place:
https://www.gov.uk/government/collections/teaching-excellence-framework

```{r read_explore_data, warning=FALSE,include=FALSE}
library(tidyverse)

#Reading in TEF_YearFourAllMetrics 
#Also reading contextual data: this is analysed in later section (see below)

webread <- 0 #set this to 1 to read direct from web
if (webread==1){
#Use these lines to read direct from web - slower
tef4file <- read_csv("https://ofslivefs.blob.core.windows.net/files/TEF%20data%20updated/Metrics/TEF_YearFour_AllMetrics.csv")
tef4context <- read_csv("https://ofslivefs.blob.core.windows.net/files/TEF%20data%20updated/Contextual/TEF_YearFour_AllContext.csv")
}

#Use these lines to read files saved to working directory - quicker
if(webread==0){
tef4file<-read_csv('rawdata//TEF_YearFour_AllMetrics.csv')
#colnames(tef4file)
tef4context<-read_csv('rawdata//TEF_YearFour_AllContext.csv')
}
```

Look at columns to see what data they contain.

```{r inspectvars, include=FALSE}
print('METRIC')
unique(tef4file$METRIC)
print('SPLIT_ID')
unique(tef4file$SPLIT_ID)
print('SPLIT_CATEGORY')
unique(tef4file$SPLIT_CATEGORY)

```

The critical 'flags' variable is coded as characters. Recode to numbers. Also make other column names less unwieldy.

```{r renamecols, include=FALSE}
#change unwieldy names
colnames(tef4file)<-c('uniname','UKPRN','majmode','has.metrics','metric','study.mode','split.id','split.cat','numerator','denom',
                      'indicator','benchmark','prcnt.b','diff','sd','zscore','flag','absval','reason')

#recode flags
tef4file$flag<-recode(tef4file$flag, "--" = 1, "-" = 2, "=" = 3, "+" = 4,"++" = 5,  .default = 9)
w<-which(tef4file$flag==9)
tef4file$flag[w]<-NA
```

# Analysis of how number of students affects ratings

The final TEF rankings will depend heavily (though not exclusively) on the 'flags' that denote how far an institution's scores depart from the expected value computed by benchmarking. 

These are computed for several different measures, including selected ratings from NSS, data on course continuation, and LEO data on employment outcomes.

The flag is determined by the joint occurence of an extreme z-score and an absolute percentage difference from the predicted value exceeding a given value. As I understand it, this was done because it was recognised that reliance on z-scores alone would make it easier for large institutions to obtain positive or negative flags, because the larger N would make the estimate more precise. 

The analysis below suggests that this problem has not been overcome.  However, I am making what sense I can of the data files provided, which are not easy to understand.  


```{r explore, include=FALSE}
#Initial plot of zscore vs demoninator : which I assume reflects N students for whom data available.
#But at this point, we have several entries for each institution.
plot(log(tef4file$denom),tef4file$zscore,xlab='Log N students (denominator)',ylab='zscore')

#plot(log(tef4file$denom),tef4file$sd) #to see how SD declines with size


```


```{r explore2, include=FALSE}
#Who are those with huge numbers?
  w<-which(tef4file$denom>15000)
unique(tef4file$uniname[w])
xcol<-c(2,3,4) #these just make it hard to view file
temp<-tef4file[w,-xcol]
#N.B. Open University is v large and has poor score on Continuation, though good on other things.
```


For the next analysis I decided to focus just on those rows specifying Core and FullTime entries: I think this means that each institution is represented just once. 

The following plots/tables look at the occurrence of extreme flags in relation to the size of the institution (as reflected in the denominator - this is plotted on a log scale)

This varies from measure to measure, but, if I have analysed this correctly, then there is a substantial increase in the proportion of flagged institutions among larger institutions for both positive and negative flags.

As well as plotting the data continuously, I have divided the institutions into quartiles by size (again using the denominator value) and computed the proportion of flags of different types for each of these quartiles. This shows the same basic thing: more extreme flags for the larger institutions (quartile 4).

There are also differences in distribution of flags across measures - perhaps reflecting distributional properties of the underlying measures?

The number of extreme flags seems remarkably high in some cases, especially for large institutions.

```{r selectcore, echo=FALSE}
tefcore <- filter(tef4file,split.id=='Core',denom>0,study.mode=='FullTime')
#Split into quartiles 1-4 in terms of size (numerator)

#do plots by each metric and work out % in each band by size quartile
tefcore$metric<-as.factor(tefcore$metric)

mycols<-c('purple','blue','grey','pink','red')
#levels(tefcore$metric)<-c('Support','Feedback','Continue','Employed','Skilled','LEOhigh','LEOsustain','Teaching')
for (i in 1:length(levels(tefcore$metric))){
  temp<-filter(tefcore,metric==levels(tefcore$metric)[i])
  plot(log(temp$denom),temp$zscore,main=levels(tefcore$metric)[i],col=mycols[temp$flag],ylim=c(-25,25),
       ylab='zscore',xlab='Log N in denominator')
  legend("bottomleft",title="Flag",c("--","-","=","+","++"), fill=mycols,horiz=TRUE)
  temp$nquant <- cut(temp$denom, quantile(temp$denom,na.rm=TRUE), include.lowest = TRUE,labels=1:4)
  mytab<-table(temp$nquant,temp$flag)
  colnames(mytab)<-c('--','-','=','+','++')
  rownames(mytab)<-c('Size.quartile1','Size.quartile2','Size.quartile3','Size.quartile4')
  print('Proportions of the 5 flag values for each size quartile')
  print(levels(tefcore$metric)[i])
  print(prop.table(mytab,1))
}


```

# Conclusion and questions

I have put these questions to tefmetrics@officeforstudents.org.uk and will update when I get a reply.

1. Have I analysed this right? Please do let me know if I have misunderstood what the data represent.

2. If these analyses are correct, should we be concerned that the larger institutions are far more likely to obtain extreme scores (flagged up) than smaller institutions on most measures?

3. Also the proportions of extreme flags appear to differ between measures. Is this to be expected? (I thought the definition of flags should avoid this, but could the combination of using z-scores with an absolute value cutoff have different consequences for different measures?)

4. N.B. I just selected cases where split.id=='Core' and study.mode=='FullTime'. I am not entirely confident I understand what is meant by 'Core' - does this just subsume  all other categories?

5. Also, it seems that in some cases the 'denominator' does not match the 'head count' in the AllContext file. Is this just a consequence of missing data? Or does the denominator in some cases span more than one year? Is any adjustment made for missing data, given that it cannot be assumed it is missing at random?

# Update: 9th February 2019
Here is the response from OfS, which is reassuring in suggesting I have interpreted the file correctly.

<i>Having looked at your analysis it seems that you have understood what the data represents and analysed in a way that makes sense of TEF methodology.
 
The issue of smaller student numbers ‘defaulting’ to silver is something we are aware of. Paragraph 94 on page 29 of the report on findings from the first subject pilot (1) mentions some OfS analysis on this. The Government consultation response (2) also has a section on this. On page 40, the government response to question 10 refers to assessability, and potential methods that could be used to deal with this in future runs of the TEF.
 
In response to your other questions, the core split is the overall split for a metric. All other splits are essentially a subset of this.
 
The denominator may not match the contextual data for a number of reasons. There are several exclusion reasons for each metric, meaning students which are included in the contextual data may not be included in the metrics themselves. The exclusion reasons for each metric can be found on page 58 onwards of the TEF Year 4 procedural guidance (3).</i>

The email response linked directly to pdfs. Relevant links to these sources are:
(1) https://www.officeforstudents.org.uk/publications/teaching-excellence-and-student-outcomes-framework-findings-from-the-first-subject-pilot-2017-18/
(2) https://www.gov.uk/government/consultations/teaching-excellence-and-student-outcomes-framework-subject-level (This has government response)
(3) https://www.officeforstudents.org.uk/publications/teaching-excellence-and-student-outcomes-framework-year-four-procedural-guidance/




# Further analysis, 28 Jan 2019


## Consideration of flags in relation to raw scores (before benchmarking)

I assume the raw scores are the percentage of students who meet the benchmark for a specific indicator.
Below I have plotted the data to show the relationship between the % meeting the indicator, and the derived z-score (which is used to determine the flags), separating the institutions according to how big they are.
(The line is just the linear regression for each institution size-quartile).

This analysis just confirms how the size of institution affects the awarding of flags. 
For large institutions, even a small change in % meeting benchmark has a big impact on flags. 
A small institution, on the other hand, won't get a flag (either positive or negative) regardless of how extreme their score is.
Now, it could be argued that this is right because the estimate for a small institution will be unreliable. But it is counterintuitive that, for instance a small institution with close to 100% on 'academic support' does not get a + or ++ flag, whereas a large institution can get ++ for a score of 80%.



```{r plotraw,echo=FALSE}
nucols<-c('orange','green','brown','black')
pchlist <-c(20,4,1,3,8) #attempt to use vaguely meaningful symbols
for (i in 1:length(levels(tefcore$metric))){
  temp<-filter(tefcore,metric==levels(tefcore$metric)[i])
    temp$nquant <- as.numeric(cut(temp$denom, quantile(temp$denom,na.rm=TRUE), include.lowest = TRUE,labels=1:4))
#  plot(log(temp$denom),temp$indicator,main=levels(tefcore$metric)[i],col=mycols[temp$flag],pch=15,
 #      ylab='% meeting Indicator',xlab='Log N in denominator')
 # legend("bottomright",title="Flag",c("--","-","=","+","++"), fill=mycols,horiz=TRUE)

  #look at how benchmarking affects flags
  plot(temp$zscore~temp$indicator,col=nucols[temp$nquant],pch=pchlist[temp$flag],
       main=paste('Raw x z-score coded by size of HEI\n',main=levels(tefcore$metric)[i]),xlab='Indicator',ylab='z-score')
# Add a legend
  legpos<-min(temp$indicator,na.rm=TRUE)
legend(legpos, 15,title='N Quartile',legend=c('Small','Med','Large','Huge'),
       col=nucols,pch=15,cex=.8)  
legend(legpos+10, 15,title="Flag",c("--","-","=","+","++"), pch=pchlist,cex=.8)
 #Add regression lines separated by Nquartile
 for (n in 1:4){
   thisfile<-filter(temp,nquant==n)
   myl<-lm(thisfile$zscore~thisfile$indicator)
   abline(a=myl$coefficients[1],b=myl$coefficients[2],col=nucols[n])
 }


  #create table of means by quartile
  group_summary <- temp %>% 
    group_by(nquant) %>% 
   summarise_at( vars(denom, indicator), c(mean, sd), na.rm=TRUE)

 
  group_summary<-(data.frame(group_summary))
  colnames(group_summary)<-c('Quartile','Mean N','Mean indicator','SD N','SD indicator')
  group_summary<-group_summary[,c(1,2,4,3,5)]
  print(group_summary)
  

}
```

# Implications

1. The difficulty of getting meaningful benchmarking raises big questions about the workability of benchmarking at subject-level, when numbers will be far smaller. 

2. The numbers with extreme flags seem too high, especially for large institutions.

A huge amount of data on a large set of variables is being subjected to intensive and complicated computation (just follow the links to various workbooks and instruction manuals for the TEF statistics to see what I mean) solely in order to end up with a 3-point ranking of institutions as gold, silver or bronze.

This involves mixing a basket of quite disparate measures, none of which rates teaching quality. 
The analyses presented here raise doubts about the validity of the statistical methods applied to the data to derive the 3-point scale.

The logic of benchmarking can also be questioned, though that raises a different set of issues. Benchmarking in effect assumes that if students from disadvantaged backgrounds do more poorly on a measure such as satisfaction with the course, we should take that into account in rating the institution. An alternative view is that institutions should adapt their courses for the students so that this association disappears.

We are told we need TEF because students want information, but they can get information from Unistats https://unistats.ac.uk/. This has information about the different variables that are input into TEF. For different students, different things matter: some may be more concerned about future earnings than others, for instance. Others may think the evaluations of prior students are more important. Unistat allows them to see how specific courses fare on these variables, as well as providing information on entry requirements. TEF does not.

# Contextual data

Additional analysis started 9th Feb 2019. 

```{r explore.context}
head(tef4context)

mysize <- filter(tef4context,CONTEXT_ID=='Overall',CONTEXT_CATEGORY=='FTE' )
hist(mysize$TOTAL_HEADCOUNT)
#v skewed! try log
hist(log(mysize$TOTAL_HEADCOUNT))

#Suggests we could omit institutions with fewer than 500 students
w<- which(mysize$TOTAL_HEADCOUNT > 500)
includelist <- mysize$PROVIDER_NAME[w] #Most of those < 500 are specialist colleges

mycontext <- filter(tef4context, PROVIDER_NAME %in% includelist)

print(paste('N institutions with at least 500 headcount:  ',length(unique(mycontext$PROVIDER_NAME))))

#select just these institutions from tefcore

tefcore2 <- filter(tefcore,uniname %in% includelist )


```


The number of students on whom data is available appears quite variable from indicator to indicator.
To look at this I need to compare the Ns in the mycontext dataframe with the denominators in tefcore2.

I'll create a wide-form data frame,starting with just the overall data from mycontext, and then add denominators from tefcore2 in additional columns.

```{r makewide}
mywide <- data.frame(filter(mycontext,CONTEXT_ID=='Overall',CONTEXT_CATEGORY=='Headcount' ))
k<-length(mywide) #N columns
#add columns for new variables
mywide$NSS <- 0
mywide$Continue <- 0
mywide$Employ<-0
mywide$Earn<-0
mywide$NSS.p <- 0
mywide$Continue.p <- 0
mywide$Employ.p<-0
mywide$Earn.p<-0

mycats <- c("TheTeachingOnMyCourse","Continuation","EmploymentOrFurtherStudy","LEOAboveMedianEarningsThresholdOrFurtherStudy")

for (i in 1: nrow(mywide)){
  temp <- filter(tefcore2,uniname==mywide$PROVIDER_NAME[i])
  for (j in 1:length(mycats)){
  w <- which(temp$metric==mycats[j])
  if(length(w>0)){
    mywide[i,(k+j)] <- temp$denom[w]
  }
  }
}

for (j in (k+1):(k+4)){
  mywide[,(j+4)]<-round(100*mywide[,j]/mywide$TOTAL_HEADCOUNT,1)
  hist(mywide[,(j+4)],main=colnames(mywide)[(j+4)])
  plot(log(mywide$TOTAL_HEADCOUNT),mywide[,(j+4)],main=colnames(mywide)[(j+4)],ylab="% with data")
}


```

There's something odd about the Continuation data - it seems some institutions have continuation data for more than 100% of the students in their headcount!

I wondered whether I'd misaligned the headcount values, but the distributions for the other metrics seem sensible - with values > 100% rare enough to be occasional data entry errors.

I am comparing across 2 files, so it's worth checking against some specific examples.

```{r checkNcont}
#Find those where N datapoints on continuations is > 105% relative to headcount
w<-which(mywide$Continue.p>105)
mywide$PROVIDER_NAME[w]

#Looks like these are mostly colleges of FE or new universities 
#Let's now check a few original downloaded files - picking those that are universities


mytemp <- filter(mycontext,PROVIDER_NAME=='Leeds Trinity University')
#confirms total headcount is 2835

mytemp2<-filter(tef4file,uniname=='Leeds Trinity University')
mytemp3<-filter(mytemp2,metric=='Continuation')
```

This shows for Leeds Trinity University a denominator of 3100 for full time and 46 for part time students for .
So the denominator in the TEF4yr metric data is greater than the headcount in the contextual data.

Maybe these are measured at different time points?

# Redoing my original analyses with actual headcount rather than denominator


```{r addsize}
#Need to add the headcount variable for each institution to the tefcore2 file
tefcore2$headcount <-0  #create column

for (i in 1: nrow(mywide)){
  w <- which(tefcore2$uniname==mywide$PROVIDER_NAME[i])
  tefcore2$headcount[w]<- mywide$TOTAL_HEADCOUNT[i]
  }

```

```{r plottrueN, echo=FALSE}
#Split into quartiles 1-4 in terms of size (numerator)

#do plots by each metric and work out % in each band by size quartile

mycols<-c('purple','blue','grey','pink','red')
#levels(tefcore$metric)<-c('Support','Feedback','Continue','Employed','Skilled','LEOhigh','LEOsustain','Teaching')
for (i in 1:length(levels(tefcore2$metric))){
  temp<-filter(tefcore2,metric==levels(tefcore$metric)[i])
  plot(log(temp$headcount),temp$zscore,main=levels(tefcore$metric)[i],col=mycols[temp$flag],ylim=c(-25,25),
       ylab='zscore',xlab='Log N for headcount')
  legend("bottomleft",title="Flag",c("--","-","=","+","++"), fill=mycols,horiz=TRUE)
  temp$nquant <- cut(temp$headcount, quantile(temp$headcount,na.rm=TRUE), include.lowest = TRUE,labels=1:4)
  mytab<-table(temp$nquant,temp$flag)
  colnames(mytab)<-c('--','-','=','+','++')
  rownames(mytab)<-c('Headcount.quartile1','Headcount.quartile2','Headcount.quartile3','Headcount.quartile4')
  print('Proportions of the 5 flag values for each size quartile')
  print(levels(tefcore$metric)[i])
  print(prop.table(mytab,1))
}


```

The effect of HEI size looks less striking if you take actual headcount rather than denominator.

This raises the question of the relationship between headcount and denominator


```{r true.vs.denom}


for (i in 1:length(levels(tefcore2$metric))){
  temp<-filter(tefcore2,metric==levels(tefcore$metric)[i])
  plot(temp$headcount,temp$denom,main=levels(tefcore$metric)[i],col=mycols[temp$flag],
       ylab='denominator',xlab='N for headcount')
  legend("topleft",title="Flag",c("--","-","=","+","++"), fill=mycols,horiz=TRUE)
}
```